The Name of the Title Is Hope

### 2024-12-30 paper.tex
done with my read of assignment #3

\section{Introduction}
(Tell readers we are targetting WebGPU)
(Speed of light definition here)
\subsection{Goals of a Scan Implementation}
\begin{itemize}
  \item \textbf{Minimal global memory access:} the scan should minimize global memory access to the theoretical minimum of $2n$.
  \item \textbf{Fully saturates global memory bandwidth:} the scan should be capable of fully saturating the global memory bandwidth.
  \item \textbf{Portability:} the scan should be portable across hardware vendors and architectures.
\end{itemize}

@@

In addition to its speed, the \emph{Chain-Scan} \john{Be consistent, Chained-Scan or Chain-Scan?} architecture has other, more practical properties, that improve over its predecessors. As it is a single kernel, it minimizes launch overheads and simplifies maintenance. On previous architectures, handling of arbitrary inputs \john{input sizes?} can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogenous system to coordinate them. Similarly, a raking approach \john{(a necessary optimization in approaches prior to Chained-Scan) ?} can be cumbersome, as determining the workgroup occupancy for every target device is often impractical. Furthermore, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. As a consequence of its $2n$ global data movement and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation. In previous architectures, such tasks often required additional kernel launches or complex bookkeeping to handle dependencies between workgroups. By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, minimizing overhead \john{Your audience might not have any idea what you are talking about in the next phrase. Add something like ``augmenting scan with useful extensions such as \ldots''?} and enabling seamless integration of compaction and allocation operations.

@@

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery} and at worst, risks the aforementioned TDR and subsequent program crash. This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG\@.

Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support.\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query} This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware. In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}. However, as discussed earlier, \emph{Reduce-Then-Scan} carries its own set of drawbacks, including increased global memory traffic ($O(3n)$) and additional kernel launches, which result in reduced throughput and higher latency. While these tradeoffs were once deemed acceptable, they stand in stark contrast to the efficiency and simplicity of \emph{Chained-Scan} when implemented on hardware with reliable FPG\@. Yet, in the absence of guaranteed forward progress, developers have little choice but to sacrifice performance for broader compatibility.

\subsection{Earlier Attempts at Portability (Cuttable?)}
Levien and Naur~\cite{}, made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress in \emph{DecoupledLookback}, introducing \emph{ScallarFallback}. \john{Scalar, not Scallar?} In \emph{ScallarFallback}, each time a workgroup waits on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time. Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can also perform the reduction of a blocking tile itself, thus breaking the inter-workgroup dependency.

@@

\section{Design}
\subsection{Goals}
This work is guided by two main goals. First, portability: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and achitectures, including those without FPG\@. Second, performance: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model. To ground these objectives, we select the WebGPU shading language (WGSL) and the WebGPU standard as our target environment. WGSL is a meta-level shading language which is translated and compiled as necessary for backend graphics APIs---D3D12, Vulkan, and Metal---and as such, WGSL implementations are bound by the minimum capability across all backends. Because WGSL must operate within this limited capability space, it inherently embodies the challenge of portability.

\subsection{Non Goals}
Although our goal is to create a fully portable and highly performant scan implementation, we are constrained by underlying hardwares and programming models. As discussed in more detail in \emph{Limits on the Speed-of-Light}, not all architectures offer atomic operations that are sufficiently fast enough or scheduling models that are sufficeintly fair enough to attain speed-of-light performance, and thus we cannot guarantee such performance. Although we are cognisant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control,\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in it.} placing a solution to subgroup divergence issues outside our scope. Furthermore, although real-world scans may require more than 30 bits of space for their reductions, or may involve scanning across multiple struct members, WGSL’s current limitations---lack of fences and 64-bit atomics---lead us to focus on the simplest scenario that fits within the 30-bit range. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.

@@

This increased sensitivity to fairness exacerabtes existing issues which negatively impact the performance of previous scan architectures, namely lack of compute, and slow atomic update latency. On hardware that lacks sufficient compute power to be memory-bandwidth bound, the additional work incurred by redundant fallback reductions is particularly deleterious. High atomic update latency further compounds the problem: as inter-workgroup communication time grows, the minimum delay before updates become visible to dependent workgroups also increases. This, in turn, raises the number of lookback steps that may be needed and potentially leads to redundant fallbacks.

>>>>

\section{Introduction}
(Tell readers we are targeting WebGPU)
(Speed of light definition here)
\subsection{Goals of a Scan Implementation}
\begin{itemize}
  \item \textbf{Minimal global memory access:} the scan should minimize global memory access to the theoretical minimum of $2n$.
  \item \textbf{Fully saturates global memory bandwidth:} the scan should be capable of fully saturating the global memory bandwidth.
  \item \textbf{Portability:} the scan should be portable across hardware vendors and architectures.
\end{itemize}

@@

In addition to its speed, the \emph{Chain-Scan} \john{Be consistent, Chained-Scan or Chain-Scan?} architecture has other, more practical properties, that improve over its predecessors. As it is a single kernel, it minimizes launch overheads and simplifies maintenance. On previous architectures, handling of arbitrary inputs \john{input sizes?} can be awkward: the recursive approach, for example, entails multiple kernel launches and a heterogeneous system to coordinate them. Similarly, a raking approach \john{(a necessary optimization in approaches prior to Chained-Scan) ?} can be cumbersome, as determining the workgroup occupancy for every target device is often impractical. Furthermore, on GPUs where the global memory data cache is indexed by physical addresses—and therefore requires lookups in \emph{translation lookaside buffers} (TLBs)—raking the workgroups can lead to performance degradation due to cache thrashing. This issue occurs when task sizes are large enough that each workgroup accesses its own unique page entry, and the device lacks sufficient TLB coverage to support the memory access patterns of all workgroups. As a consequence of its $2n$ global data movement and inherently serial ordering, \emph{Chained-Scan} is also suitable for \emph{in-situ} compaction and allocation. In previous architectures, such tasks often required additional kernel launches or complex bookkeeping to handle dependencies between workgroups. By contrast, \emph{Chained-Scan} efficiently propagates dependencies within a single kernel, minimizing overhead \john{Your audience might not have any idea what you are talking about in the next phrase. Add something like ``augmenting scan with useful extensions such as \ldots''?} and enabling seamless integration of compaction and allocation operations.

@@

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
Although NVIDIA formalized FPG down to the intra-subgroup level beginning with the Volta architecture---and FPG was likely already present at the workgroup level on at least NVIDIA's Fermi and AMD's TeraScale2 architectures~\cite{}---it remains absent on some contemporary hardware, most notably certain ARM-based chips, including Apple's M-series GPUs~\cite{10.1145/3485508}. Our testing shows that, at best, attempting to run \emph{Chained-Scan} without FPG results in mega-bad,\footnote{Need to gather data on the distribution of M1 Pro time on Decoupled Lookback instead of the average time. It's also unclear why the test did not trigger Metal's timeout device recovery. \john{This is super cool, that you have these results. Simply showing a distribution of runtimes (that presumably is a wide distribution) and a fraction of failures will be just fine. I could see putting these results right here instead of in the results section.}} and at worst, risks the aforementioned TDR and subsequent program crash. This extreme slowdown highlights the inherent fragility of \emph{Chained-Scan} without FPG\@.

Exacerbating this issue is the fact that no major graphics API currently provides developers with a mechanism to query FPG support.\footnote{As of writing this, 12/29/2024, we are unaware of any major graphics API---Vulkan, D3D12, or Metal---that provide developers with an FPG capability query. \john{I like that you are being polite and guarded about this, but when you're 100\% sure, you can just assert it without this qualification.}} This lack of transparency forces developers to design algorithms without clear knowledge of whether critical guarantees are available on the target hardware. In environments like shading languages, where cross-platform portability is required, this uncertainty compels developers to revert to older, less efficient strategies such as \emph{Reduce-Then-Scan}. However, as discussed earlier, \emph{Reduce-Then-Scan} carries its own set of drawbacks, including increased global memory traffic ($O(3n)$) and additional kernel launches, which result in reduced throughput and higher latency. While these tradeoffs were once deemed acceptable, they stand in stark contrast to the efficiency and simplicity of \emph{Chained-Scan} when implemented on hardware with reliable FPG\@. Yet, in the absence of guaranteed forward progress, developers have little choice but to sacrifice performance for broader compatibility.

\subsection{Earlier Attempts at Portability (Cuttable?)}
\john{It's cuttable only in the sense that it's not peer-reviewed literature (right?) but I think it is appropriate to cite it and this description you have made is a good one. Besides showing the problem doesn't have a simple solution, it also totally underscores that you have done your homework to find the relevant prior scan work. (As a reviewer, I make a lot of judgments based on the quality of a background section; an author can make me favorably inclined toward their work by showing they really understand, and can place in proper context, the prior work on the topic. Bonus points for citing work I don't know.)}
Levien and Naur~\cite{} made the first attempt at implementing a \emph{DecoupledLookback} approach without forward progress in \emph{DecoupledLookback}, introducing \emph{ScallarFallback}. \john{Scalar, not Scallar?} In \emph{ScallarFallback}, each time a workgroup waits on a preceding tile, it begins a \emph{fallback} operation, and begins consuming the blocking tile one element at a time. Although this entails performing redundant work when the blocking tile is simply late but not deadlocking, it guarantees the termination of the algorithm because a workgroup can also perform the reduction of a blocking tile itself, thus breaking the inter-workgroup dependency.

@@

\section{Design}

\john{I think here you want a Problem Statement. What are you actually building? Here is where you say ``scan of 30-bit ints, arbitrary size, arbitrary monoid'', that sort of thing. Below in ``Non Goals'' you make reference to 30 bits and structs, but you haven't formally stated what you're building. Here's where you state this. It'll also be in the abstract.}

\subsection{Goals}
This work is guided by two main goals:

\begin{enumerate}
  \item \textbf{Portability}: we want to develop a scan implementation that retains the benefits of the \emph{Chained-Scan} architecture but is also suitable for a diverse range of hardware vendors and architectures, including those without FPG\@.
  \item \textbf{Performance}: we aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.
\end{enumerate}

\noindent
To ground these objectives, we select the WebGPU shading language (WGSL) and the WebGPU standard as our target environment. WGSL is a meta-level shading language which is translated and compiled as necessary for backend graphics APIs---D3D12, Vulkan, and Metal---and as such, WGSL implementations are bound by the minimum capability across all backends. Because WGSL must operate within this limited capability space, it inherently embodies the challenge of portability.

\subsection{Non Goals}
Although our goal is to create a fully portable and highly performant scan implementation, we are constrained by underlying hardware platforms and programming models. As discussed in more detail in \emph{Limits on the Speed-of-Light}, not all architectures offer atomic operations that are sufficiently fast enough or scheduling models that are sufficiently fair enough to attain speed-of-light performance, and thus we cannot guarantee such performance. Although we are cognizant of the risks posed by subgroup divergence, subgroup functions in shading languages do not allow explicit divergence control,\footnote{For example, CUDA allows developers to explicitly provision subgroup functions with a mask of the threads that will participate in it.} placing a solution to subgroup divergence issues outside our scope. Furthermore, although real-world scans may require more than 30 bits of space for their reductions, or may involve scanning across multiple struct members, WGSL’s current limitations---lack of fences and 64-bit atomics---lead us to focus on the simplest scenario that fits within the 30-bit range. Lastly, we do not investigate alternative $O(2n)$ scan architectures beyond \emph{Chained-Scan}.

\john{So now, actually talk about the design. Let's say you wanted to present to, say, Guy Blelloch, who knows scans inside and out but doesn't know much about GPUs. How would you explain the approach of decoupled fallback, without diving into GPU details?}

@@

This increased sensitivity to fairness exacerbates existing issues which negatively impact the performance of previous scan architectures, namely lack of compute, and slow atomic update latency. On hardware that lacks sufficient compute power to be memory-bandwidth bound, the additional work incurred by redundant fallback reductions is particularly deleterious. High atomic update latency further compounds the problem: as inter-workgroup communication time grows, the minimum delay before updates become visible to dependent workgroups also increases. This, in turn, raises the number of lookback steps that may be needed and potentially leads to redundant fallbacks.
